---
author: "Ruthger Righart"
date: "16.02.2017"
output:
  html_document: default
  pdf_document: default
---

# <span style="color:black">R programming | Using R for Exploratory Analyses</span>

## <span style="color:orange"> ANALYSES AND R PROGRAMMING CODE

### Create a small dataset: 100m sprint

```{r, echo=TRUE, message=FALSE, eval=TRUE}
df<-data.frame(subject=as.factor(c("Pierre","Anne","Joyce","Peter","Alan","Camille")), age=as.numeric(c(20, 16, 19, 99, 23, 18)), sex=as.factor(c("M","F","F","M",NA,"F")), height=as.numeric(c(172, 181, 165, 168, 177, 178)), speed=as.numeric(c(11.20,3.00,11.50,10.35,10.98,13.05)))
```

Check the data.frame df by simply typing in:

```{r, echo=TRUE}
df
```

The data has 6 rows and 5 columns:
```{r, echo=TRUE}
dim(df)
```

Because the data are not that big you are able to view them entirely in your R studio. However, if you have very big data you may want to view only the first lines:
```{r, echo=TRUE}
head(df, 4)
```

To inspect the structure of the data: 
```{r, echo=TRUE}
str(df)
```

Some basic statistics can be given by the summary command:

```{r, echo=TRUE}
summary(df)
```

### Missing values

To examine if missing values (NA) are in the data:
```{r, echo=TRUE}
is.na(df)
```

If you want the row and column index where NAs occur:
```{r, echo=TRUE}
which(is.na(df), arr.ind=T)
```

### Impossible extreme values
We can use boxplots to see if there are impossible extreme values:

```{r, echo=TRUE}
par(mfrow=c(1,3))
boxplot(df$age, main="Age (yrs)", cex.lab=2.0, cex.axis=2.0, cex.main=1.6, cex=2.0, col="yellow")
boxplot(df$height, main="Height (cm)", cex.lab=2.0, cex.axis=2.0, cex.main=1.6, cex=2.0, col="red")
boxplot(df$speed, main="Speed (ms)", cex.lab=2.0, cex.axis=2.0, cex.main=1.6, cex=2.0, col="orange")
```

### Finding & replacing the extreme values manually

Compute the mean *before* the removal of outliers:

```{r, echo=TRUE}
mean(df$age)
```

Check if there are cases that are older than 40 years:

```{r, echo=TRUE}
df$age>40
```

Replace the case(s) older than 40 with a missing value (NA).

```{r, echo=TRUE}
df$age[df$age>40]<-NA
```

Compute again the mean age, allowing to remove missing values (NAs):
```{r, echo=TRUE}
mean(df$age, na.rm=TRUE)
```

### Inspecting mean age for male and female participants
Mean age for male and female participants. 

```{r, echo=TRUE}
aggregate(age ~ sex, data=df, FUN=mean, na.rm=TRUE)
```

### Inspecting relations between variables

```{r, echo=FALSE, message=FALSE}
df[2,5]<-NA
```


Use a scatterplot to display the relation between age and speed:
```{r, echo=TRUE}
par(mfrow=c(1,1))
plot(speed ~ age, data=df, col="blue", pch=1, cex=1.2)
```

# <span style="color:orange">QUIZ QUESTIONS AND ANSWERS</span>

### Quiz Question I

Replace the outlier of 3.00 ms in the variable *df$speed* with a NA.

**Answer**
There are multiple solutions possible:

```{r, echo=TRUE}
df$speed[df$speed==3.00]<-NA
```

Another solution:

```{r, echo=TRUE}
df[2,5]<-NA
```

To verify that the value was indeed replaced by a NA use the command `is.na`:

```{r, echo=TRUE}
is.na(df$speed)
```

### Quiz Question II

A dataset could erroneously have double records (duplicates). This is bad and should be removed. 

Seeing the current dataset, what would be a way to discover duplicates in the variable *df$subject*?

Check your solution for the following data.frame that contains double data:

```{r, echo=TRUE, message=FALSE, eval=TRUE}
df<-data.frame(subject=as.factor(c("Pierre","Anne","Joyce","Peter","Alan","Camille", "Pierre")), age=as.numeric(c(20, 16, 19, 99, 23, 18, 20)), sex=as.factor(c("M","F","F","M",NA,"F", "M")), height=as.numeric(c(172, 181, 165, 168, 177, 178, 172)), speed=as.numeric(c(11.20,3.00,11.50,10.35,10.98,13.05, 11.20)))
```

**Answer**

A way to inspect for double records is to use the `table` function.

```{r, echo=TRUE, message=FALSE, eval=TRUE}
table(df$subject)
```

But still simpler, using the `summary` function would also display this:

```{r, echo=TRUE, message=FALSE, eval=TRUE}
summary(df)
```



